---
title: "`r params$project_name`"
subtitle: "Bulk RNA-seq analysis report"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
params:
  project_name: "Bulk RNA-seq analysis"
  outdir: "results"
  scripts_dir: "."
  design: "condition"
  contrasts: ""
  alpha: 0.05
  lfc_threshold: 0
  top_n: 40
  quantifier: "salmon"
  run_de: true
  de_problems: ""
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
    theme: flatly
  pdf_document:
    toc: true
    toc_depth: 2
    latex_engine: tectonic
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 4.8, dpi = 150)
suppressPackageStartupMessages({
    library(ggplot2)
    library(knitr)
})
outdir <- params$outdir
alpha <- as.numeric(params$alpha)
lfc_thresh <- as.numeric(params$lfc_threshold)
top_n <- as.integer(params$top_n)
run_de <- isTRUE(params$run_de) || identical(params$run_de, "TRUE")
contrast_names <- if (nzchar(params$contrasts)) strsplit(params$contrasts, ",")[[1]] else character(0)

PALETTE <- c("#4C72B0", "#DD8452", "#55A868", "#C44E52", "#8172B3",
             "#937860", "#DA8BC3", "#8C8C8C", "#CCB974", "#64B5CD")
rtheme <- function() {
    theme_bw(base_size = 11) +
        theme(panel.grid.minor = element_blank(),
              strip.background = element_rect(fill = "grey92", colour = NA),
              plot.title = element_text(face = "bold", size = 12))
}
note_missing <- function(what) {
    cat("\n*", what, "was not available for this run - this section is omitted.*\n\n")
}
# Point labels: use ggrepel when it happens to be installed, otherwise a nudged
# geom_text. Keeps ggrepel an optional rather than a required dependency.
point_labels <- function() {
    if (requireNamespace("ggrepel", quietly = TRUE)) {
        ggrepel::geom_text_repel(aes(label = sample), size = 3, show.legend = FALSE,
                                 max.overlaps = 20)
    } else {
        geom_text(aes(label = sample), size = 3, vjust = -0.9, show.legend = FALSE)
    }
}
read_tsv_safe <- function(path, ...) {
    if (!file.exists(path)) return(NULL)
    out <- tryCatch(read.delim(path, check.names = FALSE, ...), error = function(e) NULL)
    if (is.null(out) || nrow(out) == 0) return(NULL)
    out
}

counts  <- read_tsv_safe(file.path(outdir, "03_quant", "gene_counts.tsv"), row.names = 1)
tpm     <- read_tsv_safe(file.path(outdir, "03_quant", "gene_tpm.tsv"), row.names = 1)
vst     <- read_tsv_safe(file.path(outdir, "04_de", "vst.tsv"), row.names = 1)
coldata <- read_tsv_safe(file.path(outdir, "04_de", "coldata.tsv"), row.names = 1)

# Salmon writes per-sample mapping statistics to aux_info/meta_info.json
read_salmon_meta <- function(outdir) {
    d <- file.path(outdir, "03_quant", "salmon")
    if (!dir.exists(d)) return(NULL)
    samples <- list.dirs(d, recursive = FALSE, full.names = FALSE)
    rows <- lapply(samples, function(s) {
        f <- file.path(d, s, "aux_info", "meta_info.json")
        if (!file.exists(f)) return(NULL)
        txt <- paste(readLines(f, warn = FALSE), collapse = " ")
        num <- function(key) {
            m <- regmatches(txt, regexpr(paste0('"', key, '"\\s*:\\s*[0-9.eE+-]+'), txt))
            if (length(m) == 0) return(NA_real_)
            suppressWarnings(as.numeric(sub('.*:\\s*', '', m)))
        }
        lib <- regmatches(txt, regexpr('"library_types"\\s*:\\s*\\[\\s*"[^"]*"', txt))
        data.frame(sample = s,
                   num_processed = num("num_processed"),
                   num_mapped = num("num_mapped"),
                   percent_mapped = num("percent_mapped"),
                   library_type = if (length(lib)) sub('.*"([^"]*)"$', "\\1", lib) else NA_character_,
                   stringsAsFactors = FALSE)
    })
    rows <- Filter(Negate(is.null), rows)
    if (length(rows) == 0) return(NULL)
    do.call(rbind, rows)
}
salmon_meta <- read_salmon_meta(outdir)

summary_dir <- file.path(outdir, "05_report", "summary_tables")
dir.create(summary_dir, recursive = TRUE, showWarnings = FALSE)
write_summary <- function(df, name) {
    if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
    write.table(df, file.path(summary_dir, paste0(name, ".tsv")),
                sep = "\t", quote = FALSE, row.names = FALSE)
}
```

# Overview

```{r overview}
ov <- data.frame(
    Item = c("Samples", "Genes quantified", "Quantification",
             "Differential expression", "Design", "Contrasts",
             "Significance threshold"),
    Value = c(
        if (!is.null(counts)) ncol(counts) else NA,
        if (!is.null(counts)) nrow(counts) else NA,
        if (params$quantifier == "salmon") "Salmon (selective alignment)" else "STAR + Salmon",
        if (run_de) "DESeq2" else "not run",
        if (run_de) paste0("~", params$design) else "-",
        if (length(contrast_names)) paste(contrast_names, collapse = ", ") else "-",
        sprintf("adjusted p < %g%s", alpha,
                if (lfc_thresh > 0) sprintf(", |log2FC| >= %g", lfc_thresh) else "")
    ), stringsAsFactors = FALSE)
ov <- ov[!is.na(ov$Value) & ov$Value != "-", ]
kable(ov, caption = "Analysis summary")
write_summary(ov, "run_overview")
```

`r if (!run_de && nzchar(params$de_problems)) paste0("**Differential expression was not run.** ", params$de_problems, "\n\nQuantification and QC results below are complete and usable; add or correct the design/condition columns in the sample sheet to enable DE.")`

Gene-level counts were produced with
`r if (params$quantifier == "salmon") "Salmon selective alignment" else "STAR genomic alignment followed by Salmon quantification"`
and aggregated with tximport using length-scaled TPM, which corrects for both library
size and transcript-length differences between samples.

# Quality control

## Mapping rates

```{r mapping}
if (is.null(salmon_meta)) {
    note_missing("Salmon mapping statistics")
} else {
    m <- salmon_meta
    m$percent_mapped <- round(m$percent_mapped, 2)
    write_summary(m, "mapping_rates")
    kable(m[order(m$percent_mapped), ],
          row.names = FALSE, digits = 2,
          col.names = c("Sample", "Reads processed", "Reads mapped", "% mapped", "Library type"),
          caption = "Salmon mapping statistics")
}
```

```{r mapping-plot, fig.height=3.6}
if (!is.null(salmon_meta)) {
    ggplot(salmon_meta, aes(x = reorder(sample, percent_mapped), y = percent_mapped)) +
        geom_col(fill = PALETTE[1]) +
        geom_hline(yintercept = 70, linetype = "dashed", colour = "grey35") +
        coord_flip() +
        labs(title = "Mapping rate by sample",
             subtitle = "dashed line = 70%, a common minimum for usable bulk RNA-seq",
             x = NULL, y = "% reads mapped") +
        rtheme()
}
```

`r if (!is.null(salmon_meta)) "Samples mapping well below the rest of the batch usually indicate contamination, the wrong reference, or degraded input material. Treat expression estimates from such samples cautiously."`

## Library size and gene detection

```{r libsize, fig.height=3.6}
if (is.null(counts)) {
    note_missing("Count matrix")
} else {
    ls_df <- data.frame(sample = colnames(counts),
                        total_counts = colSums(counts),
                        genes_detected = colSums(counts > 0),
                        stringsAsFactors = FALSE)
    write_summary(ls_df, "library_sizes")
    kable(ls_df, row.names = FALSE, digits = 0,
          col.names = c("Sample", "Total counts", "Genes detected"),
          caption = "Library size and number of detected genes")
    print(
        ggplot(ls_df, aes(x = reorder(sample, total_counts), y = total_counts / 1e6)) +
            geom_col(fill = PALETTE[3]) + coord_flip() +
            labs(title = "Library size", x = NULL, y = "Counts (millions)") + rtheme()
    )
}
```

# Sample relationships

```{r pca, fig.height=4.6}
mat <- if (!is.null(vst)) as.matrix(vst) else if (!is.null(tpm)) log2(as.matrix(tpm) + 1) else NULL
if (is.null(mat) || ncol(mat) < 2) {
    note_missing("Expression matrix for ordination")
} else {
    vars <- apply(mat, 1, var)
    keep <- head(order(vars, decreasing = TRUE), min(500, sum(vars > 0)))
    if (length(keep) < 2) {
        note_missing("Enough variable genes for PCA")
    } else {
        pca <- prcomp(t(mat[keep, , drop = FALSE]), scale. = FALSE)
        pv <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)
        pdat <- data.frame(sample = rownames(pca$x),
                           PC1 = pca$x[, 1],
                           PC2 = if (ncol(pca$x) > 1) pca$x[, 2] else 0,
                           stringsAsFactors = FALSE)
        grp_col <- NULL
        if (!is.null(coldata)) {
            cand <- intersect(strsplit(params$design, "[ +~*:]+")[[1]], colnames(coldata))
            cand <- cand[nzchar(cand)]
            if (length(cand)) {
                grp_col <- cand[1]
                pdat[[grp_col]] <- coldata[pdat$sample, grp_col]
            }
        }
        write_summary(pdat, "pca_coordinates")
        p <- ggplot(pdat, aes(x = PC1, y = PC2,
                              colour = if (!is.null(grp_col)) .data[[grp_col]] else NULL))
        print(
            p + geom_point(size = 3.2) +
                point_labels() +
                scale_colour_manual(values = PALETTE, na.value = "grey50",
                                    name = grp_col) +
                labs(title = "Principal component analysis",
                     subtitle = "top 500 most variable genes",
                     x = sprintf("PC1 (%.1f%%)", pv[1]),
                     y = sprintf("PC2 (%.1f%%)", if (length(pv) > 1) pv[2] else 0)) +
                rtheme()
        )
    }
}
```

```{r corr-heatmap, fig.height=5}
if (!is.null(mat) && ncol(mat) >= 2) {
    cm <- cor(mat, method = "spearman")
    cdf <- expand.grid(x = rownames(cm), y = colnames(cm), stringsAsFactors = FALSE)
    cdf$r <- as.vector(cm)
    print(
        ggplot(cdf, aes(x = x, y = y, fill = r)) +
            geom_tile() +
            geom_text(aes(label = sprintf("%.2f", r)), size = 2.6, colour = "grey15") +
            scale_fill_gradient(low = "#F7F7F7", high = PALETTE[1], name = "Spearman r") +
            labs(title = "Sample-to-sample correlation", x = NULL, y = NULL) +
            rtheme() +
            theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
    write_summary(data.frame(sample = rownames(cm), cm, check.names = FALSE),
                  "sample_correlation")
}
```

`r if (!is.null(mat)) "Samples from the same experimental group are expected to cluster together. A sample sitting with the wrong group, or far from everything else, is worth investigating before interpreting the differential expression results."`

# Differential expression

```{r de-setup}
de_tables <- list()
if (run_de && length(contrast_names)) {
    for (cn in contrast_names) {
        f <- file.path(outdir, "04_de", cn, "results.tsv")
        d <- read_tsv_safe(f)
        if (!is.null(d)) de_tables[[cn]] <- d
    }
}
```

```{r de-summary}
if (length(de_tables) == 0) {
    note_missing("Differential expression results")
} else {
    summ <- do.call(rbind, lapply(names(de_tables), function(cn) {
        d <- de_tables[[cn]]
        sig <- !is.na(d$padj) & d$padj < alpha & abs(d$log2FoldChange) >= lfc_thresh
        data.frame(Contrast = cn,
                   Tested = sum(!is.na(d$padj)),
                   Significant = sum(sig),
                   Up = sum(sig & d$log2FoldChange > 0),
                   Down = sum(sig & d$log2FoldChange < 0),
                   stringsAsFactors = FALSE)
    }))
    write_summary(summ, "de_summary")
    kable(summ, row.names = FALSE, caption = sprintf(
        "Differentially expressed genes (adjusted p < %g%s)", alpha,
        if (lfc_thresh > 0) sprintf(", |log2FC| >= %g", lfc_thresh) else ""))
}
```

```{r de-plots, results='asis', fig.height=4.4}
for (cn in names(de_tables)) {
    d <- de_tables[[cn]]
    cat("\n\n##", cn, "\n\n")
    d$sig <- !is.na(d$padj) & d$padj < alpha & abs(d$log2FoldChange) >= lfc_thresh
    d$direction <- ifelse(!d$sig, "n.s.",
                          ifelse(d$log2FoldChange > 0, "up", "down"))

    # Volcano
    dv <- d[is.finite(d$log2FoldChange) & !is.na(d$padj), ]
    if (nrow(dv) > 0) {
        dv$neglog10p <- -log10(pmax(dv$padj, .Machine$double.xmin))
        p <- ggplot(dv, aes(x = log2FoldChange, y = neglog10p, colour = direction)) +
            geom_point(alpha = 0.6, size = 1.3) +
            scale_colour_manual(values = c("up" = PALETTE[4], "down" = PALETTE[1],
                                           "n.s." = "grey78"), name = NULL) +
            geom_hline(yintercept = -log10(alpha), linetype = "dashed", colour = "grey40") +
            labs(title = paste("Volcano plot -", cn),
                 x = "log2 fold change", y = "-log10 adjusted p") +
            rtheme()
        print(p)
    }

    # MA
    dm <- d[is.finite(d$baseMean) & d$baseMean > 0, ]
    if (nrow(dm) > 0) {
        p <- ggplot(dm, aes(x = baseMean, y = log2FoldChange, colour = direction)) +
            geom_point(alpha = 0.6, size = 1.1) +
            scale_x_log10() +
            scale_colour_manual(values = c("up" = PALETTE[4], "down" = PALETTE[1],
                                           "n.s." = "grey78"), name = NULL) +
            geom_hline(yintercept = 0, colour = "grey30") +
            labs(title = paste("MA plot -", cn),
                 x = "mean normalised count (log scale)", y = "log2 fold change") +
            rtheme()
        print(p)
    }

    # Top genes table
    dsig <- d[d$sig, ]
    dsig <- dsig[order(dsig$padj, -abs(dsig$log2FoldChange)), ]
    show <- head(dsig, top_n)
    if (nrow(show) > 0) {
        cols <- intersect(c("gene_id", "gene_name", "baseMean", "log2FoldChange", "padj"),
                          colnames(show))
        print(kable(show[, cols], row.names = FALSE, digits = c(0, 0, 1, 3, 6),
                    caption = sprintf("Top %d significant genes - %s", nrow(show), cn)))
        write_summary(show[, cols], paste0("top_genes_", gsub("[^A-Za-z0-9_.-]", "_", cn)))
    } else {
        cat("\n*No genes passed the significance threshold for this contrast.*\n\n")
    }
}
```

```{r heatmap, fig.height=6}
if (length(de_tables) > 0 && !is.null(vst)) {
    d <- de_tables[[1]]
    d <- d[!is.na(d$padj), ]
    d <- d[order(d$padj), ]
    genes <- head(d$gene_id, 30)
    genes <- intersect(genes, rownames(vst))
    if (length(genes) >= 2) {
        m <- as.matrix(vst[genes, , drop = FALSE])
        # centre each gene so the heatmap shows relative, not absolute, expression
        m <- t(scale(t(m), center = TRUE, scale = FALSE))
        labels <- genes
        if ("gene_name" %in% colnames(d)) {
            nm <- d$gene_name[match(genes, d$gene_id)]
            labels <- ifelse(is.na(nm) | nm == "", genes, nm)
        }
        hdf <- expand.grid(gene = labels, sample = colnames(m),
                           stringsAsFactors = FALSE)
        hdf$value <- as.vector(m)
        hdf$gene <- factor(hdf$gene, levels = rev(labels))
        print(
            ggplot(hdf, aes(x = sample, y = gene, fill = value)) +
                geom_tile() +
                scale_fill_gradient2(low = PALETTE[1], mid = "white", high = PALETTE[4],
                                     midpoint = 0, name = "centred\nexpression") +
                labs(title = paste("Top differential genes -", names(de_tables)[1]),
                     subtitle = "variance-stabilised, gene-centred",
                     x = NULL, y = NULL) +
                rtheme() +
                theme(axis.text.x = element_text(angle = 45, hjust = 1),
                      axis.text.y = element_text(size = 7))
        )
    }
}
```

# Methods and provenance

```{r provenance}
prov <- data.frame(
    Item = c("Report generated", "Quantification", "Aggregation",
             "Differential expression", "Output directory"),
    Value = c(format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
              if (params$quantifier == "salmon") "Salmon selective alignment"
              else "STAR alignment + Salmon quantification",
              "tximport, length-scaled TPM",
              if (run_de) sprintf("DESeq2, design ~%s", params$design) else "not run",
              normalizePath(outdir, mustWork = FALSE)),
    stringsAsFactors = FALSE)
kable(prov, row.names = FALSE, caption = "Provenance")
```

Tables behind every figure are written to `05_report/summary_tables/` as TSV, and full
per-contrast results (all genes, not only significant ones) are in
`04_de/<contrast>/results.tsv`.

**Interpretation caveats.** Adjusted p-values control the false discovery rate across
genes within a contrast, not across contrasts. Fold changes for very lowly expressed
genes are unstable even after shrinkage; the `baseMean` column indicates how much
evidence supports each gene. Statistical significance is not biological importance -
a small, reproducible change in a key regulator may matter more than a large change in
an abundant structural transcript. Batch effects that are confounded with the condition
of interest cannot be corrected by any model and must be addressed in the experimental
design.
